---
title: "R Notebook"
output: html_notebook
---


```{r 'setup', message=FALSE, include=FALSE, results='hide'}
#require("knitr")
#opts_knit$set(root.dir = "~/Desktop/David/Projects/gapminder")
options(scipen = 999)
```


Libraries for analysis

```{r 'libraries', message=FALSE, warning=FALSE}

library(data.table)
library(ggplot2)
library(gganimate)
library(stringr)
library(gapminder)

```


Load Gapminder R package data, but this starts only in 1952

```{r 'r-gapminder'}

data("gapminder")

gapminder[order(gapminder$year),]

```


Download population, income and life expenctancy data from gapminder website

```{r 'load-data'}
# Make directory
if(!dir.exists("data")) dir.create("data")

# https://www.gapminder.org/data/
#https://docs.google.com/spreadsheets/d/192pjt2vtwAQzi154LJ3Eb5RF8W9Fx3ZAiUZy-zXgyJo/pub#
urls <- c("https://docs.google.com/spreadsheet/pub?key=phAwcNAVuyj1jiMAkmq1iMg&output=csv", 
          "https://docs.google.com/spreadsheet/pub?key=phAwcNAVuyj2tPLxKvvnNPA&output=csv")

# names we give files
names <- c("income_cap",
           "life_expectancy",
           "population")

# function to download, rename and put in file
variables <- lapply(urls, fread)


download.file("http://gapm.io/dl_pop", destfile = "data/population.xlsx")
population <- setDT(readxl::read_xlsx("data/population.xlsx", sheet = 7, col_names = FALSE))
population <- population[,-c(1,3), with=TRUE]
variables <- rlist::list.append(variables, population)
names(variables) <- names

```


Load the data frame from a file into a list of data frames called "files"

```{r eval=FALSE, include=FALSE}

# Get files from "data" folder, load raw data and give names
subdir <- "./data/"
names <- str_remove(gtools::mixedsort(list.files(subdir)), ".csv")
files <- lapply(list.files(subdir, full.names = TRUE), fread)
names(files) <- names

files

```


Convert to "Tidy" format

```{r 'convert-to-tidy'}

# Add names, convert the data frames into "tidy" data and set key for new data frame called "variables"
tidy <- 
  mapply(function(variable, name) {
    setnames(variable, as.character(variable[1]))
    names(variable)[1] <- "country"
    variable <- variable[-1,]
    variable <-
      melt(variable, 
           variable.name = "year", 
           value.name = name,
           variable.factor = FALSE)
},
variables,
names,
SIMPLIFY = FALSE)

```


Combine all three variables by country and year

```{r 'join-variables'}

# "Join" tidy data frames together on the country and year keys
gm <- 
  Reduce(function(x, y) merge(x, y), tidy)

gm

#saveRDS(gm, "data/gapminder.RDS")
```


Add continent variable

```{r 'continent-key'}

# Create continent key and set key
continent_key <- 
  unique(as.data.table(gapminder)[,.(country,continent)])

setkey(continent_key, country)

continent_key

# Add continent to data
gm_full <- continent_key[gm]


```

Summarize and clean up

```{r 'summarize'}

# Summarize
skimr::skim(gm_full)

# Clean up
rm(list=(ls()[!grepl('gm_full',ls())]))
```


Plot by continent for 2017

```{r 'plot-2017'}

ggplot(gm_full[year=="2017"], 
       aes(income_cap, 
           life_expectancy_years, 
           size = population_total, 
           colour = country)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  #scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  scale_x_log10(labels=scales::dollar) +
  labs(title = 'Year', x = 'GDP per capita', y = 'Life Expectancy') +
  facet_wrap(~continent) +
  theme_bw()

```


Explore and drop "NA" rows

```{r 'missing-rows'}

# Find countries not mapped to continents
missing <- unique(gm_full[is.na(continent)][,country])
gapminder[gapminder$country %in% missing,]
missing

# Remove countries with unknown continent
gm_final <- 
  gm_full[!is.na(continent)]
```

Final plot with animation like Hans Rosling

```{r 'animation'}
# https://gganimate.com

ggplot(gm_final, 
       aes(income_cap, 
           life_expectancy_years, 
           size = population_total, 
           colour = country)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  scale_x_log10(labels=scales::dollar) +
  facet_wrap(~continent) +
  # Here comes the gganimate specific bits
  labs(title = 'Year: {frame_time}', 
       x = 'GDP per capita', 
       y = 'life expectancy') +
  transition_time(as.integer(year)) +
  ease_aes('linear') 
  #+ shadow_wake(0.20, wrap = FALSE)

  #anim_save("gapminder.gif")
```

